\section{Introduction}

Academic authors frequently write using \LaTeX -- which is a typesetting system that has been specifically developed to produce high-quality technical and scientific documents \cite{latex}. 

With \LaTeX\ there are two distinct layers of information; the first layer contains the author's actual writing content and the second layer represents the rules that govern how the content will appear in print.

Therefore, with \LaTeX\ it becomes possible for publication and editing staff to enforce and/or adhere to consistent formatting throughout a publication, while still allowing authors complete freedom as regards the content of their writing.

Although \LaTeX\ greatly facilitates the work of editorial staff who manage publications, it does require authors to learn some new terminology and conventions. 

For professional authors and those who regularly write academic papers, learning this new terminology and conventions is simply another part of the job. However, for many undergraduate students, learning these new terms and conventions is much more challenging.

Additionally, over recent years large language models (LLMs), including generative artificial intelligence (GenAI), have become increasingly effective as productivity tools across a wide range of disciplines.

However, within academic environments, the increasing reliance on GenAI also poses significant risks to the maintenance of academic integrity, primarily because GenAI may inadvertently contribute to plagiarism or result in "hallucinated" references/citations.

This report presents Spartan Write: An AI Agent for Academic Documentation. Spartan Write was created with the intention of providing AI support to academic authors in order to assist them during the process of creating their documentation, while at no time altering their content. The fundamental concept behind Spartan Write is as follows: **Spartan Write only uses content that the author has provided**. Therefore, Spartan Write has focused solely upon assisting authors with issues related to format, syntax and layout.

At its heart, Spartan Write combines an intelligent agent with \LaTeX. While \LaTeX\ maintains control over ensuring consistent formatting, the agent assists with author input through natural conversations. This may include tasks such as structuring tables, placing figures and captions in documents written in \LaTeX, handling \LaTeX\ syntax, organizing content, etc. Additionally, the intelligent agent included in Spartan Write may also be used to scan tables (from hand-written notes) and then convert them into \LaTeX. Figure~\ref{fig:authoring-workflow} illustrates the improvement proposed by Spartan Write to the traditional authoring workflow.
