\section{Related Work}

\subsection{\LaTeX\ usability}

\LaTeX\ gives the user significant power and flexibility in producing documentation in many areas of science and technology. 
Experienced \LaTeX\ users can often produce documents faster when drafting them. 
In contrast, inexperienced users of \LaTeX\ (first-time and/or one-time) cannot achieve the same efficiency. 
\LaTeX\ is particularly effective in typesetting mathematics expressions. 
To leverage this capability, Microsoft Word, Google Docs, and LibreOffice provide the option to include small embedded \LaTeX\ fragments within otherwise standard document markup to display mathematical expressions. 
Conversely, generating tables is typically a challenging task in \LaTeX, with both novices and experts experiencing difficulty because of the complex \LaTeX\ markup required to generate a table correctly.

\subsection{Automated \LaTeX\ Generation}

Because \LaTeX\ affords the author so much detailed control over what will appear in their final document, it comes at the cost of being very difficult to learn and master. 
Therefore, researchers have been working on finding ways to reduce the amount of time required to create \LaTeX\ source files. 
One such example includes automating the creation of tables. 
Kayal et al. \cite{kayal_tables_2023} created an automated system for identifying tables through Optical Character Recognition (OCR), and then rendering those tables via ResNet transformers.

Similar to this, Xia et al. \cite{xia_latexnet_2025}, demonstrated a method of converting images and tables into \LaTeX, by collecting images and tables along with their corresponding \LaTeX\ source code from ArXiv. 
They trained a convolutional neural network that included an image-multiplexer. 
Their results showed a high degree of similarity between the output of their system versus the expected \LaTeX\ code. 
Both of these systems exist in prototype form and, therefore, lack robustness to edge cases and unexpected inputs. 

OpenAI developed a tool named Prism that shares many characteristics with Spartan Write \cite{openai_prism}. 
Prism provides a means for authors to write papers using natural language. 
Prism stores all of the \LaTeX\ source code for the paper inside itself; similar to how Spartan Write stores all of the source code and state needed to compile the paper inside itself. 
However, unlike Spartan Write, Prism takes a less rigid approach toward ensuring academic integrity. 
Prism does a number of things including proof-reading, editing, conducting literature reviews, etc., all of which can modify the content of the original document. 
These practices are generally disallowed in most fields of study. 
Spartan Write attempts to find some middle ground by providing the authors with AI-assisted capabilities while still adhering to academic integrity guidelines.

There are currently several tools under development that are somewhat similar to Prism. 
One such example is an open-source version called OpenPrism \cite{github_openprism}. 
OpenPrism is designed to replicate the functionally provided by Prism using open-source tools. 
Bibby AI \cite{bibby_ai} is another example that produces code based upon natural language input. 
Autotype is yet another example that allows users to eliminate any need for \LaTeX\ code and instead opt for markdown-like syntax \cite{autype}. 
All three examples are simply copies of Prism, thus suffering from all of the same problems that Prism experiences.

\subsection{Generative AI in Academic Writing}

Generative AI has been adopted rapidly in creative works and software development. 
Students and researchers are discouraged from utilizing Generative AI in their academic work. 
Studies have shown an increase in usage of Large Language Models (LLMs) across published articles and associate this trend with decreased material quality, miscommunication in written expression, anomalous patterns, etc. \cite{kobak_delving_2025}. 
Surveys indicate that Generative AI is largely rejected by academics. \cite{kwon_survey_2025}.

A literature review synthesizes eleven studies conducted between 2021 and 2023 examining utilization of artificial intelligence in academic writing. 
The studies reported productivity gains for specific tasks (e.g., plagiarism detection and language learning support). 
However, the studies also emphasized concerns about academic integrity, accuracy, plagiarism and dependence on artificial intelligence.

A second survey aggregated data from twenty-four studies, documenting positive influences on the academic writing process for a variety of ways in which artificial intelligence assists writers -- e.g., generating ideas, improving the structural organization of content, supporting literature reviews, assisting authors with managing/analyzing data, and proof reading. 
They report the same concern as noted above about academic integrity.

PaperDebugger \cite{paperdebugger} is an extension platform developed atop Overleaf, an online LaTeX editor. 
PaperDebugger utilizes a plugin-based multiagent system to aid in common tasks associated with document editing. 
Overleaf has a large base of users evidenced by PaperDebugger's downloads available in the repository.

Overall, there seems to be a consensus among researchers that AI can be useful during the writing process. 
However, if it is implemented without violating ethical standards of the academic community it provides valuable assistance to authors/publishers alike.

\subsection{Other Tools}

When looking into other potential tools capable of solving the problem that Spartan Write solves, you can utilize "agencies" such as Cursor, Copilot, Claude or OpenAI Codex. 
However, as compared to these tools, Spartan Write has several distinguishing differences. 
Firstly, Spartan Write is specifically designed as a tool for academic writing purposes; whereas these tools are general-purpose tools for accomplishing arbitrary tasks. 
Secondly, safeguards for maintaining academic integrity can be custom-crafted iteratively. 
Thirdly, templates are curated manually to ensure consistency of produced output. 
Fourthly, since the agent is tied directly to both the compiler and delivered as a single package, it is aware of any failures/warnings generated by the compiler during compilation. 
Finally, since it utilizes a dedicated framework; updates containing fixes/new functionality can be shipped accordingly.
